---
title: All probe lambda values
---


Read in data and get FDR values for replication

```{r}
a <- read.csv("total_analysis_data.csv")
a$pnest_egcut <- 10^-a$pnest_egcut
a$pnest_fehr <- 10^-a$pnest_fehr
a$fdr_egcut <- p.adjust(a$pnest_egcut, "fdr")
a$fdr_fehr <- p.adjust(a$pnest_fehr, "fdr")
a$pnest_both <- NA
for(i in 1:nrow(a))
{
	a$pnest_both[i] <- pchisq(-2 * sum(log(c(a$pnest_fehr[i], a$pnest_egcut[i]))), df=4, low=FALSE)
}
a$fdr_both <- p.adjust(a$pnest_both, "fdr")
```

Remove cis-cis effects, how many remain?

```{r}
a <- subset(a, chr1 != chr2)
nrow(a)
```

How many replicate in both at FDR < 0.05

```{r}
b1 <- subset(a, (fdr_both < 0.05))
nrow(b1)
```

What are the lambda values of these probes

```{r}
plot(b1$lambda)
```

How many replicate in both at p < 0.05

```{r}
b2 <- subset(a, pnest_both < 0.05)
nrow(b2)
```

What are the lambda values of these probes

```{r}
plot(b2$lambda)
```

Do again but with no duplicate probes

```{r}
plot(subset(b2, !duplicated(gene))$lambda)
```

It looks like at replication significance threshold of FDR < 0.05 there is one cis-trans interaction that replicates that has lambda value close to 1, and at nominal p < 0.05 threshold there are 15.


Questions

1. What is the standard error around these lambda values?
